---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp = 0.618, collapse=TRUE) 
```

### In-class exercises
### Unit 2: Climate

***

### Exercise 1.1

Create the same time series figure with 95% confidence intervals for the Greenland ice sheet mass loss.

```{r}
# Read in data
ant_ice_loss = read.table("data/antarctica_mass_200204_202008.txt", skip=31, sep="", header = FALSE, col.names = c("decimal_date", "mass_Gt", "sigma_Gt")) # Antarctica ice mass loss
grn_ice_loss = read.table("data/greenland_mass_200204_202008.txt", skip=31, sep="", header = FALSE, col.names = c("decimal_date", "mass_Gt", "sigma_Gt")) # Greenland ice mass loss

#Create data.frame with an NA between the GRACE missions. Column names must match so it will merge with ant and grn data
data_break = data.frame(decimal_date=2018.0, mass_Gt=NA, sigma_Gt=NA)
#Add NA data point to the Greenland ice trends data frame
grn_ice_loss_with_NA = rbind(grn_ice_loss, data_break) # Merge grn_ice_loss data frame with our NA point
grn_ice_loss_with_NA = grn_ice_loss_with_NA[order(grn_ice_loss_with_NA$decimal_date),]
```

```{r}
head(grn_ice_loss_with_NA)
plot(mass_Gt ~ decimal_date, data=grn_ice_loss_with_NA, type='l') +
lines((mass_Gt+2*sigma_Gt) ~ decimal_date, data=grn_ice_loss_with_NA, type='l', lty='dashed') +
lines((mass_Gt-2*sigma_Gt) ~ decimal_date, data=grn_ice_loss_with_NA, type='l', lty='dashed') 
```

***

### Exercise 1.2

Calculate the average annual ice loss for each ice sheet by dividing the change in ice lost from the beginning to the end of the time series by the total time that passed. Then display the ice loss rates in a bar graph. Save the bar graph into the `figures/` directory in this repo.

```{r}
# Calculate ice loss rate (Gt / year) for Antarctica:
n_rows_ant = dim(ant_ice_loss)[1]  # Can use  this to find the last row in the dataset
delta_ice_ant = ant_ice_loss$mass_Gt[n_rows_ant] - ant_ice_loss$mass_Gt[1]
delta_date_ant = ant_ice_loss$decimal_date[n_rows_ant] - ant_ice_loss$decimal_date[1]
ant_loss_rate_Gt_per_yr = -1 * delta_ice_ant / delta_date_ant

# Calculate ice loss rate (Gt / year) for Greenland:
n_rows_grn = dim(grn_ice_loss)[1]
delta_ice_grn = grn_ice_loss$mass_Gt[n_rows_grn] - grn_ice_loss$mass_Gt[1]
delta_date_grn = grn_ice_loss$decimal_date[n_rows_grn] - grn_ice_loss$decimal_date[1]
grn_loss_rate_Gt_per_yr = -1 * delta_ice_grn / delta_date_grn

# Plot ice loss rates in a bar graph:
barplot(height=c(ant_loss_rate_Gt_per_yr, grn_loss_rate_Gt_per_yr), names.arg=c("Antarctica","Greenland"), ylab="Ice loss rate (Gt/year)")

```

To save our bar plot:

```{r}
pdf('figures/ice_loss_rate.pdf', width=5, height=5)
barplot(height=c(ant_loss_rate_Gt_per_yr, grn_loss_rate_Gt_per_yr), names.arg=c("Antarctica","Greenland"), ylab="Ice loss rate (Gt/year)")
dev.off()
```

***

### Exercise 2.1

Create a vector with the names of the oceans you have personally visited. Use the `%in%` operator to subset the oceans you have visited from the world_oceans data frame. Now use subsetting to find out whether you have ever personally visited any oceans that have a smaller area than the Atlantic ocean

```{r}
# Subset the oceans I have visited
my_ocean_names = c("Atlantic", "Pacific")
my_oceans = world_oceans[world_oceans$ocean %in% my_oceans, ]  

# Which of "my" oceans are smaller than the Atlantic?
Atlantic_area = world_oceans$area[world_oceans$ocean=="Atlantic"]
my_oceans[my_oceans$area < Atlantic_area, ]
```

***

#### Exercise 2.2

Here are a few random variables. Try converting these sentences into tests written with R code. 

```{r}
w = 15
x = c(-1, 0, 1)
y = "February"
z = c("Monday", "Tuesday", "January")
```

-  Is w greater than 10 and less than 20?
-  Are any of the values in x positive?
-  Are all of the values in x positive?
-  Is object y the word February?
-  How many values in z are days of the week?

```{r}
10 < w & w < 20
any(x > 0)
all(x > 0)
y == "February"
sum(z %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", 
  "Saturday", "Sunday"))
```

***

#### Exercise 3.1

Let's say you took someone's temperature and you want to first evaluate whether or not they have a fever. If they DO have a fever, claculate how different their temperature is from the median human temperature of 98.6 and print out the answer. Also, if they DO have a fever, evaluate whether the fever is high > 101 and if it is, print out a warning message. This will require a set of nested `if()` statements. Test your code with a temperature that is not a fever (`temp = 98.4`), that is a low fever (`temp = 99.5`) and that is a high fever (`temp = 102.1`).

```{r}
temp = 103

temp_norm = 98.6
temp_high = 101
if (temp > temp_norm){
  diff = temp - temp_norm
  print(paste("Your temp is ", round(diff, digits = 1), "*F above normal"))
  if (temp > temp_high){
    print("Go to the doctor!")
  }
}
```

***

### Exercise 3.2

Let's say you have the number of donuts your Dad bought on Sunday stored in variable `n_donuts`. Write an if/else chain to print out a statement that states whether Dad bought less than a dozen, a dozen (12), a baker's dozen (13) or more than a baker's dozen. 

```{r}
n_donuts = 5

if (n_donuts < 12){
  print("I need more donuts!!")
}else if (n_donuts == 12){
  print("Yum! A dozen donuts!")
}else if (n_donuts == 13){
  print("Excellent, a baker's dozen :)")
}else {
  print("Whoa, you bought more than a baker's dozen")
}
```

***

### Exercise 3.3

Perhaps in a dataset that you are using, the number `-999` is used to indicate a missing value (this is super common). To take advantage of R's intuitive handling of missing values (in math equations, plotting, etc.) you should replace those elements with `NA`. Use the ifelse() statement to replace all of the `-999` values in the `data` vector with `NA`. Then use an if/else chain to print out a statement about whether there are any `NA`s in the `data` variable. It may help to use functions we covered earlier like `any()` and `is.na()`.

```{r}
data = c(1,3,5,-999,2,-999,8)

data_with_nas = ifelse(data == -999, NA, data)
data_with_nas

if (any(is.na(data_with_nas))){
  print("There are NAs in this data!")
}else{
  "There are no NAs in this data"
}
```

***



